\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-09-26 parts-grn.tex
start updating sect 5 notations

\paragraph{Discussion on interval parameters}
Whereas most of the literature use single value parameters (e.g., $K_{a,\omega} = i$), we introduce
in this paper intervals as parameter values (e.g., $K_{a,\omega} = \segm{i}{j}$) for specifying
the discrete functions for components.
These interval parameters are mainly introduced for convenience, and it is worth noticing that they
do not add extra expressivity for modeling multivalued networks in the general case.
However, interval parameters are more expressive than Thomas' parameters with single value.
First, we remark that if  $K_{a,\omega} = \segm{i}{j}$ with $i < j$, then, necessarily there exists
a positive edge from $a$ to $a$ in the interaction graph of the dynamics.
Then, let us consider the case where $K_{a,\omega} = \segm{i}{i+2}$ (interval of three values), and
let $s$ be a state such that $\GRNres{a}{s} = \omega$.
From \pref{def:dynamics}, if $\GRNget{s}{a} \in K_{a,\omega}$, then $a$ cannot evolve in $s$;
therefore, $a$ has three (local) fixed points ($i$, $i+1$ and $i+2$).
This behavior cannot be reproduced with single value parameters, because it is impossible to distinguish three different sets of resources corresponding to the three involved states.
At most two steady-states can be modeled by using a self-regulation $\GRNedge{a}{s}{t}{a}$, allowing to distinguish two cases ($\GRNget{s}{a} < t$ and $\GRNget{s}{a} \geq t$), which is not enough.

>>>>

\todo{(LP) parameters equivalence:
if $\GRNedge{a}{+}{t}{a}$,
$K_{a,R\setminus \{a\}} = t
\equiv
K_{a,R\setminus \{a\}} = u$
for $t \leq u \leq l_a$;
$K_{a,R\cup \{a\}} = t-1
\equiv
K_{a,R\cup \{a\}} = u$
for $0 \leq u < t$.
If $\GRNedge{a}{-}{t}{a}$,
$K_{a,R\cup \{a\}} = t
\equiv
K_{a,R\cup \{a\}} = u$
for $t \leq u \leq l_a$;
$K_{a,R\setminus \{a\}} = t-1
\equiv
K_{a,R\cup \{a\}} = u$
for $0 \setminus u < t$.
- any other equivalence?}

### 2014-09-26 parts-infer-param.tex
TODO on parameters equivalence
forgot to highlight my changes
tr2global++
start updating sect 5 notations

\todo{In this while section: add informal explanations and maybe more examples (see previous section).}

Given the IG inferred from a PH as presented in the previous section, one can find the discrete parameters
that model the behavior of the studied PH using the method presented in the following.
It relies on an exhaustive enumeration of all predecessors of each component in order to find attractor processes
and returns a possibly incomplete parametrization, given the exhaustiveness of the cooperations (\pref{ssec:infer-K}).
The last step consists of the enumeration of all compatible complete parametrizations (\pref{ssec:admissible-K}) given this
set of inferred parameters, the PH dynamics and some biological constraints on parameters.

@@

This subsection presents some results related to the inference of independent discrete parameters from a given PH.
These results are equivalent to those presented in \cite{PMR10-TCSB}, with notation adapted to be shared with the previous section.
In addition, we introduce the well-formed PH for parameter inference property (\pref{pro:wf-ph-K}),
which implies that %the inferred IG does not contain any unsigned interactions, and thus can be seen as the
%regular IG $(\Gamma, E)$,
%and that 
any process in $\levels{b}{a}$ (resp. $\ulevels{b}{a}$) share the same behavior
regarding $a$.

@@

\begin{property}[Well-formed PH for parameter inference]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for IG inference, and
the IG $(\Gamma, E)$ inferred by \pref{pps:inference-IG}
verifies the following property:
\todo{Define $c(\sigma)$ for $\sigma \in \textstyle\prod_{d \in \PHpredec{c}} \PHl_d$
(\ie the unique focal point)}
\begin{align*}
  \begin{split}
%   \forall a \in \Gamma&, \forall b\in \GRNreg{a},
%           \forall (i,j\in\levels{b}{a} \vee i,j\in\ulevels{b}{a}), \\
%   & \quad \forall c \in \PHs, ( (b\neq a\wedge c=a) \vee (c\in\PHpredec{a}\setminus \Gamma \wedge b\in\PHdirectpredec{c})), \\
%   & \qquad
%                           \PHfrappe{b_i}{c_k}{c_l}\in\PHa \Leftrightarrow
%                                   \PHfrappe{b_j}{c_k}{c_l}\in\PHa
      &
    \forall a \in \Gamma,
    \forall g \in X(a),
    \forall \sigma, \sigma' \in \textstyle\prod_{d \in g} \PHl_d,
    \forall b \in g,
    b \in \GRNreg{c},
    \forall c \in \PHdirectpredec{a} \setminus \Gamma,
      \\&\qquad
    b \in \reg{c},
    \forall (i,j \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}),
%    \forall b_i \in \PHl_b,
    b_i \in \sigma,
%    b_i \in \levels{b}{a},
%    \forall b_j \in \levels{b}{a},
      \\&\qquad\qquad
    \sigma' = \sigma\{b_j\},
    %c_{\sigma} \in \PHl_c \wedge c_{\sigma'} \in \PHl_c,  %facultatif
    \PHfrappe{c(\sigma)}{a_i}{a_j} \in \PHa \Leftrightarrow
    \PHfrappe{c(\sigma')}{a_i}{a_j} \in \PHa
  \end{split}\\
  \begin{split}
    & \forall a \in \Gamma, \forall b \in \GRNreg{a},
      \forall (i,j \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}), \\
    & \qquad \forall c \in \PHs, b \neq a,
      \PHfrappe{b_i}{a_k}{a_l} \in \PHa \Leftrightarrow
      \PHfrappe{b_j}{a_k}{a_l} \in \PHa
  \end{split}
\end{align*}
\end{property}

@@

For each sort $b \in \GRNreg{a}$, we define a context $C^b_{a,\omega}$ in \pref{eq:param_context} that contains all processes representing the influence of the resources in the configuration modeled by $\omega$.
The context of a cooperative sort $\upsilon$ that regulates $a$ is given in
\pref{eq:param_context_coop} as the set of focal processes matching the current configuration.
$C_{a,\omega}$ refers to the union of all these contexts (\pref{eq:K-ctx}).
\begin{align}
  \label{eq:param_context}
  \forall b\in\Gamma,~
  C_{a,\omega}^b & \DEF \begin{cases}
    \levels{b}{a} & \text{if $b \in \omega$,}\\
    \ulevels{b}{a} & \text{if $b \notin \omega$,}\\
    L_b            & \text{otherwise;}\\
  \end{cases}
  \\
  \label{eq:param_context_coop}
  \forall \upsilon \in \PHpredec{a}\setminus\Gamma,~
  C_{a,\omega}^\upsilon & \DEF \{
  \upsilon(\sigma) \mid \sigma \in \textstyle\prod_{c\in\PHdirectpredec{\upsilon}}C_{a,\omega}^c \}
  \\
  C_{a,\omega} & \DEF \textstyle\bigcup_{b\in\PHpredec{a}} C^b_{a,\omega}
  \label{eq:K-ctx}
\end{align}

The parameter $K_{a,\omega}$ specifies to which values $a$ eventually evolves as long as the context
$C_{a,\omega}$ holds, which is precisely the definition of the $\focals$ function
(\pref{def:focals} in \pref{ssec:focal}),
where the focals reachability property can be derived from \pref{pro:wf-ph-K} and
\pref{eq:param_context_coop}.
Hence $K_{a,\omega} = \focals(a,C^a_{a,\omega},C_{a,\omega})$ if this latter is a non-empty interval
(\pref{pps:param_K}).

\begin{proposition}[Parameter inference]
\label{pps:param_K}
Let $(\PHs, \PHl, \PHh)$ be a Process Hitting well-formed for parameter inference, and $\IG = (\Gamma, E)$ the inferred IG.
Let $A$ (resp. $B$) $\subseteq \Gamma$ be the set of regulators that activate (resp. inhibit) a sort
$a$.
If $\focals(a,C^a_{a,\omega},C_{a,\omega})=\segm{a_i}{a_j}$ is a non-empty interval, then $K_{a,\omega} = \segm{i}{j}$.
\end{proposition}

@@

\begin{property}[Parameter validity]\label{pro:K-valid}
A parameter $K_{a,\omega}$ is valid w.r.t. the PH if and only if the following equation is verified:
\begin{align*}
  \forall a_i\in C^a_{a,\omega}, a_i \notin K_{a,\omega} \Longrightarrow
    (& \exists \PHfrappe{c_k}{a_i}{a_j}\in\PHa, c_k \in C^c_{a,\omega} \\
     & \wedge a_i < K_{a,\omega} \Rightarrow j > i \wedge  a_i > K_{a,\omega} \Rightarrow j <i )
\end{align*}
\end{property}

>>>>

\todo{update text: IG can be given by the user + details}
Given the IG inferred from a PH as presented in the previous section, one can find the discrete parameters
that model the behavior of the studied PH using the method presented in the following.
It relies on an exhaustive enumeration of all predecessors of each component in order to find attractor processes
and returns a possibly incomplete parametrization, given the exhaustiveness of the cooperations (\pref{ssec:infer-K})
\todo{not clear}.
The last step consists of the enumeration of all compatible complete parametrizations (\pref{ssec:admissible-K}) given this
set of inferred parameters, the PH dynamics and some biological constraints on parameters.
\todo{not clear}

@@

This subsection presents some results related to the inference of independent discrete parameters from a given PH.
These results are equivalent to those presented in \cite{PMR10-TCSB}, with notation adapted to be shared with the previous section.
In addition, we characterize the well-formed PH for parameter inference property (\pref{pro:wf-ph-K}),
which implies that %the inferred IG does not contain any unsigned interactions, and thus can be seen as the
%regular IG $(\Gamma, E)$,
%and that 
any process in $\levels{b}{a}$ (resp. $\ulevels{b}{a}$) share the same behavior
regarding $a$.

@@

\begin{property}[Well-formed PH for parameter inference]\label{pro:wf-ph-K}
A PH is well-formed for parameter inference if and only if
it is well-formed for BRN identification (\pref{pro:wf-ph}) and
that the associated IG $(\Gamma, E)$ verifies the following property:
\modLP{
\begin{align*}
\forall b\rightarrow a\in E&, b\neq a,
	\forall \sigma\in L(\PHpredecgene{a}),\\
    \forall (i,j & \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}),
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\})))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\})))
\end{align*}
\begin{align*}
\forall a\in\Gamma, &\forall b\in\PHpredecgene{a},
b\neq a \wedge b\rightarrow a\notin E,\\
	\forall \sigma&\in L(\PHpredecgene{a}), 
    \forall i,j \in L_b,
	\forall a_k \in L_a,
	\\
&	\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_i\})))
	=\focals(a,\{a_k\},\ctx_a^{\PHpredecgene{a}}(\sigma\{b_j\})))
\end{align*}
}
\todo{Define $c(\sigma)$ for $\sigma \in \textstyle\prod_{d \in \PHpredec{c}} \PHl_d$
(\ie the unique focal point)}
\modMF{
\begin{align*}
  \begin{split}
%   \forall a \in \Gamma&, \forall b\in \GRNreg{a},
%           \forall (i,j\in\levels{b}{a} \vee i,j\in\ulevels{b}{a}), \\
%   & \quad \forall c \in \PHs, ( (b\neq a\wedge c=a) \vee (c\in\PHpredec{a}\setminus \Gamma \wedge b\in\PHdirectpredec{c})), \\
%   & \qquad
%                           \PHfrappe{b_i}{c_k}{c_l}\in\PHa \Leftrightarrow
%                                   \PHfrappe{b_j}{c_k}{c_l}\in\PHa
      &
    \forall a \in \Gamma,
    \forall g \in X(a),
    \forall \sigma, \sigma' \in \textstyle\prod_{d \in g} \PHl_d,
    \forall b \in g,
    b \in \GRNreg{c},
    \forall c \in \PHdirectpredec{a} \setminus \Gamma,
      \\&\qquad
    b \in \reg{c},
    \forall (i,j \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}),
%    \forall b_i \in \PHl_b,
    b_i \in \sigma,
%    b_i \in \levels{b}{a},
%    \forall b_j \in \levels{b}{a},
      \\&\qquad\qquad
    \sigma' = \sigma\{b_j\},
    %c_{\sigma} \in \PHl_c \wedge c_{\sigma'} \in \PHl_c,  %facultatif
    \PHfrappe{c(\sigma)}{a_i}{a_j} \in \PHa \Leftrightarrow
    \PHfrappe{c(\sigma')}{a_i}{a_j} \in \PHa
  \end{split}\\
  \begin{split}
    & \forall a \in \Gamma, \forall b \in \GRNreg{a},
      \forall (i,j \in \levels{b}{a} \vee i,j \in \ulevels{b}{a}), \\
    & \qquad \forall c \in \PHs, b \neq a,
      \PHfrappe{b_i}{a_k}{a_l} \in \PHa \Leftrightarrow
      \PHfrappe{b_j}{a_k}{a_l} \in \PHa
  \end{split}
\end{align*}
}
\end{property}

@@

For each sort $b \in \GRNreg{a}$, we define a context $C^b_{a,\omega}$ in \pref{eq:param_context} that contains all processes representing the influence of the resources in the configuration modeled by $\omega$.
The context of a cooperative sort $\upsilon$ that regulates $a$ is given in
\pref{eq:param_context_coop} as the set of focal processes matching the current configuration.
$C_{a,\omega}$ refers to the union of all these contexts (\pref{eq:K-ctx}).
\begin{align}
  \label{eq:param_context}
  \forall b\rightarrow a\in E,~
  \sigma_{a,\omega}^b & \DEF \begin{cases}
    \modLP{\min}(\levels{b}{a}) & \text{if $b \in \omega$,}\\
    \modLP{\min}(\ulevels{b}{a}) & \text{if $b \notin \omega$}
  \end{cases}
  \\
  \modLP{
  \forall b\in\PHpredecgene a, b\rightarrow a\notin E,~
  \sigma_{a,\omega}^b} & \modLP{\DEF b_0}
  \\
  \sigma_{a,\omega} & \DEF
  \modLP{\ctx_a^{\PHpredecgene{a}}(
  	\state{\sigma^b_{a,\omega} \mid  b\in\PHpredecgene{a}})}
  \label{eq:K-ctx}
  \\
  \modLP{
  \forall a\in\Gamma,~
  S_a(\ctx)} &\modLP{\DEF
	\begin{cases}
	\{a_i\} & \text{if }a_i\in \ctx\\
	L_a & \text{otherwise.}
	\end{cases}}
\end{align}

The parameter $K_{a,\omega}$ specifies to which values $a$ eventually evolves as long as the context
$C_{a,\omega}$ holds, which is precisely the definition of the $\focals$ function
(\pref{def:focals} in \pref{ssec:focal}),
where the focals reachability property can be derived from \pref{pro:wf-ph-K} and
\pref{eq:param_context_coop}.
Hence \modLP{$K_{a,\omega} = \focals(a,S_a(\sigma_{a,\omega}),\sigma_{a,\omega})$ if this latter is
a singleton, denoting a deterministic behaviour (\pref{pps:param_K})}.

\begin{proposition}[Parameter inference]
\label{pps:param_K}
Let $(\PHs, \PHl, \PHh)$ be a Process Hitting well-formed for parameter inference, and $\IG = (\Gamma, E)$ the inferred IG.
Let $A$ (resp. $B$) $\subseteq \Gamma$ be the set of regulators that activate (resp. inhibit) a sort
$a$.
If $\focals(a,S_a(\sigma_{a,\omega}),\sigma_{a,\omega})= \{ a_i \}$, then \modLP{$K_{a,\omega} = i$}.
\end{proposition}

@@

\todo{update notations}
\begin{property}[Parameter validity]\label{pro:K-valid}
A parameter $K_{a,\omega}$ is valid w.r.t. the PH if and only if the following equation is verified:
\begin{align*}
  \forall a_i\in C^a_{a,\omega}, a_i \notin K_{a,\omega} \Longrightarrow
    (& \exists \PHfrappe{c_k}{a_i}{a_j}\in\PHa, c_k \in C^c_{a,\omega} \\
     & \wedge a_i < K_{a,\omega} \Rightarrow j > i \wedge  a_i > K_{a,\omega} \Rightarrow j <i )
\end{align*}
\end{property}
